### 2019-06-24 Memoria/Chapters/ConstrainedClustering.tex
Terminado capitulo 3

Having introduced the basic concepts of clustering, the aim of this chapter is to define the concepts related to constrained clustering, as well as to give examples of its application and highlight the benefits and problems presented by its use. To this end we take as our main reference the survey presented in \cite{davidson2007survey}.

@@

As we discussed in Chapter \ref{ch:IntroClustering}, unsupervised clustering methods are useful for structuring data referring to a particular area. An example of this can be found in text classification; In \cite{cohn2003semi} a problem proposed by Yahoo! is faced. The problem consist of, given a large number of text documents, grouping them according to a taxonomy in which documents with similar topics are nearby. For this, unsupervised clustering methods are useful, as the information on the problem initially available is limited. However, in \cite{wagstaff2001constrained} is shown that applying unsupervised clustering to certain problems, such as grouping GPS data in such a way that clusters define the lanes of a road, does not produce significant results, as the clusters obtained are far from the elongated shape that would be expected as a result. To tackle the problem, they introduced a new element in clustering, the instance-level constraints, which made it possible to include knowledge about the clusters that would guide the clustering methods to obtain the expected results. It was enough to indicate that the lanes of the road on which the vehicles circulate are four meters wide, and therefore any vehicle that is at a distance of more than 4 meters from another, perpendicular to the direction of displacement, must be located in a different cluster.

We are now in a new scenario: it is possible to incorporate additional information to the clustering process, in addition to that contained in the data set itself, to guide it in the formation of the partition and obtain more precise results. This places constrained clustering in the framework of semi-supervised learning, unlike traditional clustering methods which fall within the area of unsupervised clustering.

@@

Although they may seem simple, constraints defined as above have some very interesting properties. Must-link constraints are an example of an equivalence relationship, and therefore are symmetrical, reflexive and transitive, formalizing:

\begin{observation}
	\textbf{Must-link constraints are transitive.} Let $CC_i$ and $CC_j$ be connected components (completely connected subgraphs by \acs{ML} constraints), and let $x_i$ and $x_j$ be the instances in $CC_i$ and $CC_j$ respectively. Then $C_=(x,y): x \in CC_i, y \in CC_j \rightarrow C_=(a,b) \forall a,b: a\in CC_i, b \in CC_j$. \cite{davidson2007survey}
\end{observation}

\begin{observation}
	\textbf{Cannot-link constraints can be entailed.} Let $CC_i$ and $CC_j$ be connected components (completely connected subgraphs by \acs{CL} constraints) and let $x_i$ and $x_j$ be the instances in $CC_i$ and $CC_j$ respectively. Then $C_{\neq}(x,y): x \in CC_i, y \in CC_j \rightarrow _{\neq}(a,b) \forall a,b: a\in CC_i, b \in CC_j$. \cite{davidson2007survey}
\end{observation}

A clear example of constraint use can be found in clustering applications where there are distance measurement limitations, as in the case of GPS data. Thus, if we want that instances forming two clusters to be separated by a distance greater or equal to $\delta$, it is enough to set \acs{ML} constraints between all instances separated by less than $\delta$ units. Similarly, if we want the diameter of the clusters to be at most $\epsilon$, we must set \acs{CL} constraint between instances separated by more than $\epsilon$ units. Figure \ref{fig:DistanceConstraints} shows a graphical representation of these two types of constraints.

@@

While supervised learning involves knowing the label associated with each instance, \acf{SSL} has only a subset of tagged instances. On the other hand, in many domains the available information refers to relations between instances, and not to the specific class to which they belong. Moreover, in interactive clustering systems, a user who is not expert in the domain of the problem will probably be able to provide information in the form of constraints such as \acf{ML} and \acf{CL}. \cite{cohn2003semi}\cite{davidson2007hierarchical}, rather than providing information on what particular class certain instances belong to.

Typically, constraints are incorporated into clustering problems in two ways. They can be used to alter the instance clusters assignment rule of the method in hand, so that the solution satisfies as many constraints as possible. Alternatively, it is possible to train the distance function used by the method based on the constraints, either before or during the application of the method. In any case, the initialization phase can take the constraints into account, so that instances associated with constraints \acf{ML} will be placed in the same cluster, and those between which there is a constraint \acf{CL} will be placed in different clusters. Based on this distinction, we identify two ways of approaching the problem, those based on constraints (\textit{constraint-based}), and those based on distances (\textit{distance-based}).

@@

In constraint-based methods, the clustering method itself is modified in such a way that the available information is used to bias the search and obtain an appropriate data partition.

Regarding the degree to which the constraints have to be satisfied, we can make a distinction between the concepts of hard \cite{wagstaff2001constrained}\cite{davidson2005agglomerative} and soft \cite{law2004clustering}\cite{basu2004active}\cite{segal2003discovering}\cite{davidson2005clustering}\cite{law2005model} constraints. Hard constraints must necessarily be satisfied in the output partition of any algorithm that makes use of them, while soft constraints are taken as a strong guide for the algorithm that uses them but can be partially satisfied in the output partition \cite{seret2014new}. For the purposes of this work, we will employ the latter. There are several techniques to obtain a partition based on constraints:

@@

	\item  Modify the objective function to include a penalty term for breaking constraints \cite{demiriz1999semi} \cite{davidson2005clustering}.

	\item Group instances with additional information obtained from a conditional distribution into an auxiliary space \cite{sinkkonen2000semisupervised}.

@@

Figure \ref{fig:ConstOverDataset} shows a data set together with its associated constraints. Figure \ref{fig:ClusteringSatAllConst} proposes a possible clustering that satisfies all constraints.

\begin{figure}[bth]
	\myfloatalign
	{\includegraphics[width=.6\linewidth]{gfx/ConstClust/InputInstancesAndConst1}
	\caption[Constraints on a dataset.]{Constraints on a dataset. \cite{davidson2007survey}} \label{fig:ConstOverDataset}
	}
	{\includegraphics[width=.6\linewidth]{gfx/ConstClust/ClusteringSatAll}
	\caption[Clustering satisfying all constraints.]{Clustering satisfying all constraints. \cite{davidson2007survey}} \label{fig:ClusteringSatAllConst}
	}
\end{figure}

@@

In distance-based approaches, classic clustering methods that use a distance measurement are employed, so that the distance measurement is modified to incorporate the constraints. In this context, satisfying constraints means that instances related to constraints \acf{ML} are placed together in the space, and those related by \acf{CL} are separated.

Figure \ref{fig:MetricLearned} shows a possible clustering based on a metric learned from the constraints specified in Figure \ref{fig:ConstOverDataset2}. It should be noted that in Figure \ref{fig:MetricLearned} the space in which the data is located has been compressed on the vertical axis and widened on the horizontal axis to match the learned distance metric.

\begin{figure}[bth]
	\myfloatalign
	{\includegraphics[width=.6\linewidth]{gfx/ConstClust/InputInstancesAndConst2}
	\caption[Constraints on a dataset.]{Constraints on a dataset. \cite{davidson2007survey}} \label{fig:ConstOverDataset2}
	}
	{\includegraphics[width=.6\linewidth]{gfx/ConstClust/MetricaAprendida}
	\caption[Clustering based on a metric learned based on constraints.]{Clustering based on a metric learned based on constraints. \cite{davidson2007survey}} \label{fig:MetricLearned}
	}
\end{figure}

@@

\begin{figure}[bth]
	\myfloatalign
	\subfloat[Profile faces.]
	{\includegraphics[width=.3\linewidth]{gfx/ConstClust/AnalisisImagenes/Caras1}} \quad
	\subfloat[Front faces.]
	{\includegraphics[width=.3\linewidth]{gfx/ConstClust/AnalisisImagenes/Caras3}} \quad
	\subfloat[Faces up.]
	{\includegraphics[width=.3\linewidth]{gfx/ConstClust/AnalisisImagenes/Caras2}} \quad
	\caption[CMU database faces.]{CMU database faces. \cite{davidson2007survey}}\label{fig:CMUFacesDatabase}
\end{figure}

The method used to obtain the constraints is one of the most popular in the literature: set the number of clusters of the resulting partition equal to the number of classes in the database, and generate the constraints from a subset of tagged instances; that is, if two instances have different labels, set a \acf{CL} constraint between them, otherwise one of \acf{ML} type. Thus, between the images shown in Figure \ref{fig:FacesDatabaseCL} \acf{CL} constraints are set, since, although they belong to the same person, they do not have the same orientation.

\begin{figure}[bth]
	\myfloatalign
	{\includegraphics[width=.35\linewidth]{gfx/ConstClust/AnalisisImagenes/CarasDifOr1}} \quad
	{\includegraphics[width=.35\linewidth]{gfx/ConstClust/AnalisisImagenes/CarasDifOr2}}
	\caption[\acs{CL} constraints between faces of the same person.]{\acs{CL} constraints between faces of the same person. \cite{davidson2007survey}}\label{fig:FacesDatabaseCL}
\end{figure}

@@

\begin{figure}[bth]
	\myfloatalign
	\subfloat[Imagen original]
	{\includegraphics[width=.3\linewidth]{gfx/ConstClust/AnalisisImagenes/Aibo1}} \quad
	\subfloat[Clustering sin restricciones]
	{\includegraphics[width=.3\linewidth]{gfx/ConstClust/AnalisisImagenes/Aibo2}} \quad
	\subfloat[Clustering con restricciones]
	{\includegraphics[width=.3\linewidth]{gfx/ConstClust/AnalisisImagenes/Aibo3}} \quad
	\caption[Clustering method used in Aibo Robot navigation system.]{Clustering method used in Aibo Robot navigation system. \cite{davidson2007survey}\cite{davidson2005clustering}}\label{fig:AiboRobotClustSys}
\end{figure}

@@

Video databases are one example where constraints can be generated directly from the data domain, especially if space-time data information is available \cite{yan2006discriminative}. In time sequenced data it is possible to set \acf{ML} constraints between groups of pixels of frames close in time. This is especially useful when the task is to perform object recognition based on clustering and segmentation. It is also possible to add \acf{CL} constraints to image segment in the same snapshot, as there is a low probability that they are associated with the same object after segmentation is done. In fact, in the video analysis domain problems there are a variety of constraint extraction methods \cite{yan2006discriminative}, Figure \ref{fig:ConstExtractionVideo} shows some examples.

\begin{figure}[bth]
	\myfloatalign
	\subfloat[]
	{\includegraphics[width=.4\linewidth]{gfx/ConstClust/Videos/VideoA}}
	\quad
	\subfloat[]
	{\includegraphics[width=.225\linewidth]{gfx/ConstClust/Videos/VideoB}} \quad
	\subfloat[]
	{\includegraphics[width=.4\linewidth]{gfx/ConstClust/Videos/VideoC}}
	\quad
	\subfloat[]
	{\includegraphics[width=.4\linewidth]{gfx/ConstClust/Videos/VideoD}}
	\caption[Constraints extraction in video data.]{Constraints extraction in video data. \cite{yan2006discriminative} \cite{davidson2007survey}}\label{fig:ConstExtractionVideo}
\end{figure}

In Figure \ref{fig:ConstExtractionVideo}, image (a) corresponds to constraints extracted from tracking a person over a period of time, (b) corresponds to spatial constraints associating two objects located in the same frame, image (c) corresponds to restrictions obtained through facial recognition and (d) to those provided by the user.

@@

In gene clustering based on micro-array data, genes are represented by their expression profile in different experiments and grouped using different methods, in this case constrained clustering methods. Figure \ref{fig:GeneticApp} shows an example: these are \acf{ML} constraints set between genes based on co-occurrence data stored in the protein interaction database, which contains information about which genes (and their associated proteins) are present in the same cell processes \cite{xenarios2001dip}. This information can be used to improve results provided by clustering methods. \cite{segal2003discovering}.

\begin{figure}[!h]
	\centering
	\includegraphics[scale=0.3]{gfx/ConstClust/Genetica/Genes} 
	\caption[Gene clustering based on micro-array data.]{Gene clustering based on micro-array data. \cite{davidson2007survey}}\label{fig:GeneticApp}
\end{figure}

@@

In content classification tasks, the goal is to automatically split large amounts of documents into groups or clusters. In this case it is possible to extract constraints from multiple auxiliary resources. For example, if two documents are in the same directory, a \acf{ML} constraints could be set between them. This way it is possible to adjust the resulting clustering to suit a particular criterion, such as creating a hierarchy of documents similar to the way they are organized in the input directories structure.

@@

El clustering con restricciones resulta de gran utilidad en el procesamiento de datos de búsqueda en páginas web. Aquí, el objetivo es agrupar, de manera automática, los resultados de una consulta ambigua en el motor de búsqueda en clusters de \acs{URL}s, que se refieran al concepto introducido como consulta en diferentes contextos. En este ámbito es posible extraer las restricciones a partir de búsquedas realizadas anteriormente por los usuarios, de manera que se establece una restricción de tipo \acf{ML} entre \acs{URL}s visitadas en la misma sesión de usuario. Aplicar clustering utilizando estas restricciones puede ayudar a sesgar el resultado de las búsquedas hacia las preferencias del usuario.

\subsection{Aplicaciones en datos de audio}

Constrained clustering is very useful in the processing of search data on web pages. Here, the aim is to automatically group the results of an ambiguous query in the search engine into clusters of \acs{URL}s, which refer to the concept introduced as a query in different contexts. In this area it is possible to extract constraints from searches previously performed by users, so that a \acf{ML} constraint is set between visited \acs{URL}s in the same user session. Clustering using these constraints can help biasing search results towards user preferences.

\subsection{Aplicaciones en datos de GPS} \label{sec:GPSApp}

@@

\begin{figure}[!h]
	\centering
	\includegraphics[scale=0.3]{gfx/ConstClust/GPS/Coches} 
	\caption[GPS data use.]{GPS data use. \cite{davidson2007survey} \cite{wagstaff2001constrained}}\label{fig:GPSClustData}
\end{figure}

In this domain, real clusters have an elongated shape on the horizontal axis and are aligned perpendicularly to the motion direction. To get the resulting clusters to have this shape, we can make use of the constraints. We will set a \acf{CL} constraint between those instances more than 4 meters perpendicularly to the motion direction (since the lanes have a maximum width of 4 meters), and \acf{ML} constraints between those instances that present continuity in the motion direction axis, since it is likely that the vehicles they represent are in the same lane. This clustering model has proven to be very useful in real-time navigation \cite{wagstaff2001constrained}, allowing the user to be notified when to change lanes, or when not to leave them.

\begin{figure}[!h]
	\centering
	\includegraphics[scale=0.32]{gfx/ConstClust/GPS/Instancias} 
	\caption[Resulting Clustering from GPS data with no constraints.]{Resulting Clustering from GPS data with no constraints. \cite{davidson2007survey} \cite{wagstaff2001constrained}}\label{fig:figure15}
\end{figure}

@@

The named constrained clustering applications are the most studied in the literature, but it is worth noting that is has been applied to many other fields in recent times. The following is a list of some of those applications: advanced robotics applications \cite{davidson2005clustering, semnani2016constrained}, applied marketing \cite{seret2014new}, obstructive sleep apnea analysis \cite{mai2018evolutionary}, handwritten digits classification \cite{li2015scalable}, Internet traffic classification \cite{wang2014internet}, electoral district designing, \cite{brieden2017constrained}, among others. \textcolor{red}{Poner mas ejemplos de aplicación recientes}

@@

Given $X = \{x_1 \cdots x_u\}$, a large set of unlabeled instances, and $L = \{(x_{u+1}, y_{u+1}) \cdots (x_{u+l}, y_{u+l})\}$, a small set of labeled instances, it is common to choose two elements from $L$ (with replacement) and set a \acs{ML} constraint between them if they belong to the same class, or a \acs{CL} constraints otherwise. An appropriate method to evaluate the results provided by a clustering method is to measure the level of accuracy of this when predicting dataset $X$ labels. This normally requires specifying the number of desired clusters equal to the number of known classes in $X$ ($k = k^*$). Methods such as the RandIndex \cite{rand1971objective} or its variants, which will be discussed later, are used to measure the degree of agreement between two given partitions (clusterings).

In \cite{wagstaff2000clustering}, where constraints are generated as described above, it is shown that, when an average of the accuracy of the predictions obtained with constrained clustering algorithms is made, varying the latter between experiments, results are obtained up to 20\% better than with classical clustering techniques.

@@

	\textbf{Using constraints increases accuracy on average.}
	The performance of a method in predicting labels increases when averaged using numerous different constraint sets. \cite{davidson2007survey}
	\label{ob:AccuracyIncrease}

@@

This rule, however, is not true in all cases, as in data sets such as \textit{Tic-Tac-Toe Endgame}, no increase in predictions is achieved regardless of the number of constraints used. A possible explanation for these exceptions is based on the fact that setting $k = k^*$ is not appropriate in these cases.

@@

Although, as we have seen, the incorporation of constraints to clustering methods brings benefits in some applications, there are two main drawbacks that are discussed below, as well as possible solutions to them.

@@

Introducing constraints in clustering changes the problem it solves, which becomes: \textit{Find the best partition that satisfies all constraints}. This way, if constraints are not properly specified or if the extraction methods are inadequate, we can find that the constraints contradict each other, which means that there is no partition that satisfies them all. For instance, there is no partition that satisfies the constraints $C_=(x_i,x_j)$ and $C_{\neq}(x_i,x_j)$, regardless of the value of $k$. The same happens for $k = 2$, and the constraints $C_{\neq}(x_i, x_j)$, $C_{\neq}(x_j, x_q)$ and $CL(x_i, x_q)$. Definition \ref{def:FeasibilityProblem} formalizes the feasibility problem for non-hierarchical instance-level constrained clustering:

@@

	\textbf{Feasibility problem for non-hierarchical instance-level constrained clustering:} Given a dataset $X$, a constraint set $CS$, and the bounds on the number of clusters $k_l \leq k \leq k_u$, does there exist a partition $C$ of $X$ with $k$ clusters such that all constraints in $CS$ are satisfied? \cite{davidson2005clustering} \cite{davidson2007survey}

@@

	\textbf{Constrained clustering with \acs{CL} is analogous to the graph coloring problem.} \cite{davidson2007survey}

>>>>

Once the basics concepts about clustering have been introduced, we can focus im defining the concepts related to constrained clustering, as well as to give examples of its application and highlight the benefits and problems presented by its use. To this end we take as our main reference the survey presented in \cite{davidson2007survey}.

@@

As we discussed in Chapter \ref{ch:IntroClustering}, unsupervised clustering methods are useful for structuring data referring to a particular area. An example of this can be found in text classification; In \cite{cohn2003semi} a problem proposed by Yahoo! is addressed. The problem consist of, given a large number of text documents, grouping them according to a taxonomy in which documents with similar topics are conceptually nearby. To solve this problem, unsupervised clustering methods are useful, as the information on the problem initially available is limited. However, in \cite{wagstaff2001constrained} is shown that applying unsupervised clustering to certain problems, such as grouping \acf{GPS} data in such a way that clusters define the lanes of a road, does not produce significant results, as the clusters obtained are far from the elongated shape that would be expected as a result. To tackle the problem, they introduced a new element in clustering, the constraints, which made it possible to include domain knowledge that would guide the clustering methods to obtain the expected results. It was enough to indicate that the lanes of the road on which the vehicles circulate are four meters wide, and therefore any vehicle that is at a distance of more than 4 meters from another, perpendicular to the motion direction, must be located in a different cluster.

We are now in a new scenario: it is possible to incorporate additional information to the clustering process, in addition to that contained in the dataset itself, to guide it in the formation of the output partition and obtain more accurate results. This new approach is known as constrained clustering, a \acf{SSL} paradigm, unlike traditional clustering methods which fall within the area of unsupervised learning.

@@

Although they may seem simple, constraints defined as above have some very interesting properties. \acs{ML} constraints are an example of an equivalence relationship, and therefore are symmetrical, reflexive and transitive. Observation \ref{ob:MLConstraintsRelation} formalizes this concept. On the other hand, and although \acs{CL} do not constitute an equivalence relation, they feature other properties, formalized in Observation \ref{ob:CLConstraintsRelation}.

\begin{observation}
	\textbf{Must-link constraints are transitive.} Let $CC_i$ and $CC_j$ be connected components (completely connected subgraphs by \acs{ML} constraints), and let $x_i$ and $x_j$ be the instances in $CC_i$ and $CC_j$ respectively. Then $C_=(x_i,x_j): x_i \in CC_i, x_j \in CC_j \rightarrow C_=(x_a,x_b) \forall x_a, x_b: x_a\in CC_i, x_b \in CC_j$ \cite{davidson2007survey}.
	\label{ob:MLConstraintsRelation}
\end{observation}

\begin{observation}
	\textbf{Cannot-link constraints can be entailed.} Let $CC_i$ and $CC_j$ be connected components (completely connected subgraphs by \acs{CL} constraints) and let $x_i$ and $x_j$ be the instances in $CC_i$ and $CC_j$ respectively. Then $C_{\neq}(x_i, x_j): x_i \in CC_i, x_j \in CC_j \rightarrow _{\neq}(x_a, x_b) \forall x_a, x_b: x_a\in CC_i, x_b \in CC_j$ \cite{davidson2007survey}
	\label{ob:CLConstraintsRelation}.
\end{observation}

A clear example of constraint use can be found in clustering applications where there are distance measurement limitations, as in the case of \acs{GPS} data. Thus, if we want that instances forming two clusters to be separated by a distance greater or equal to $\delta$, it is enough to set \acs{ML} constraints between all instances separated by less than $\delta$ units. Similarly, if we want the diameter of the clusters to be at most $\epsilon$, we must set \acs{CL} constraint between instances separated by more than $\epsilon$ units. Figure \ref{fig:DistanceConstraints} shows a graphical representation of these two types of constraints.

@@

While supervised learning involves knowing the label associated with each instance, \acs{SSL} has only a subset of labeled instances. On the other hand, in many domains the available information refers to relations between instances, and not to the specific class to which they belong. Moreover, in interactive clustering systems, a user who is not expert in the problem domain will probably be able to provide information in the form of constraints such as \acs{ML} and \acs{CL}, rather than providing information on what particular class certain instances belong to \cite{cohn2003semi,davidson2007hierarchical}.

Typically, constraints are incorporated into clustering problems in two ways. They can be used to alter the instance clusters assignment rule of the method in hand, so that the solution satisfies as many constraints as possible. Alternatively, it is possible to train the distance function used by the method based on the constraints, either before or during the application of the method. In any case, the initialization phase can take the constraints into account, so that instances associated with constraints \acs{ML} will be placed in the same cluster, and those between which there is a constraint \acs{CL} will be placed in different clusters. Based on this distinction, we identify two ways of approaching the problem, those based on explicit constraints (\textit{constraints-based}), and those based on distances (\textit{distance-based}).

@@

In constraints-based methods, the clustering method itself is modified in such a way that the available information is used to bias the search and obtain an appropriate data partition.

Regarding the degree to which the constraints have to be satisfied, we can make a distinction between the concepts of hard \cite{wagstaff2001constrained,davidson2005agglomerative} and soft \cite{law2004clustering,basu2004active,segal2003discovering,davidson2005clustering,law2005model} constraints. Hard constraints must necessarily be satisfied in the output partition of any algorithm that makes use of them, while soft constraints are taken as a strong guide for the algorithm that uses them but can be partially satisfied in the output partition \cite{seret2014new}. For the purposes of this work, we will employ the latter. There are several techniques to obtain a partition based on constraints:

@@

	\item  Modifying the objective function to include a penalty term for breaking constraints \cite{demiriz1999semi,davidson2005clustering}.

	\item Grouping instances with additional information obtained from a conditional distribution into an auxiliary space \cite{sinkkonen2000semisupervised}.

@@

Figure \ref{fig:ConstOverDataset} shows a dataset together with its associated constraints. Figure \ref{fig:ClusteringSatAllConst} proposes a possible clustering that satisfies all constraints.

\begin{figure}[bth]
	\myfloatalign
	{\includegraphics[width=.6\linewidth]{gfx/ConstClust/InputInstancesAndConst1}
	\caption[Constraints on a dataset.]{Constraints on a dataset \cite{davidson2007survey}.} \label{fig:ConstOverDataset}
	}
	{\includegraphics[width=.6\linewidth]{gfx/ConstClust/ClusteringSatAll}
	\caption[Clustering satisfying all constraints.]{Clustering satisfying all constraints \cite{davidson2007survey}.} \label{fig:ClusteringSatAllConst}
	}
\end{figure}

@@

In distance-based approaches, classic clustering methods that use a distance measurement are employed, so that the distance measurement is modified to incorporate the constraints. In this context, satisfying constraints means that instances related to constraints \acs{ML} are placed together in the space, and those related by \acs{CL} are separated.

Figure \ref{fig:MetricLearned} shows a possible clustering based on a metric learned from the constraints specified in Figure \ref{fig:ConstOverDataset2}. It should be noted that in Figure \ref{fig:MetricLearned} the data space has been compressed on the vertical axis and widened on the horizontal axis to match the learned distance metric.

\begin{figure}[bth]
	\myfloatalign
	{\includegraphics[width=.6\linewidth]{gfx/ConstClust/InputInstancesAndConst2}
	\caption[Constraints on a dataset.]{Constraints on a dataset \cite{davidson2007survey}.} \label{fig:ConstOverDataset2}
	}
	{\includegraphics[width=.6\linewidth]{gfx/ConstClust/MetricaAprendida}
	\caption[Clustering based on a metric learned based on constraints.]{Clustering based on a metric learned based on constraints \cite{davidson2007survey}.} \label{fig:MetricLearned}
	}
\end{figure}

@@

\begin{figure}[bth]
	\myfloatalign
	\subfloat[Profile faces.]
	{\includegraphics[width=.3\linewidth]{gfx/ConstClust/AnalisisImagenes/Caras1}} \quad
	\subfloat[Front faces.]
	{\includegraphics[width=.3\linewidth]{gfx/ConstClust/AnalisisImagenes/Caras3}} \quad
	\subfloat[Faces up.]
	{\includegraphics[width=.3\linewidth]{gfx/ConstClust/AnalisisImagenes/Caras2}} \quad
	\caption[CMU faces database.]{CMU faces database \cite{davidson2007survey}.}\label{fig:CMUFacesDatabase}
\end{figure}

The method used to obtain the constraints is one of the most popular in the literature: set the number of clusters of the resulting partition equal to the number of classes in the database, and generate the constraints from a subset of labeled instances; that is, if two instances have different labels, set a \acs{CL} constraint between them, otherwise one of \acs{ML} type. Thus, between the images shown in Figure \ref{fig:FacesDatabaseCL} \acs{CL} constraints are set, since, although they belong to the same person, they do not have the same orientation.

\begin{figure}[bth]
	\myfloatalign
	{\includegraphics[width=.35\linewidth]{gfx/ConstClust/AnalisisImagenes/CarasDifOr1}} \quad
	{\includegraphics[width=.35\linewidth]{gfx/ConstClust/AnalisisImagenes/CarasDifOr2}}
	\caption[\acs{CL} constraints between faces of the same person.]{\acs{CL} constraints between faces of the same person \cite{davidson2007survey}.}\label{fig:FacesDatabaseCL}
\end{figure}

@@

\begin{figure}[bth]
	\myfloatalign
	\subfloat[Imagen original]
	{\includegraphics[width=.3\linewidth]{gfx/ConstClust/AnalisisImagenes/Aibo1}} \quad
	\subfloat[Clustering sin restricciones]
	{\includegraphics[width=.3\linewidth]{gfx/ConstClust/AnalisisImagenes/Aibo2}} \quad
	\subfloat[Clustering con restricciones]
	{\includegraphics[width=.3\linewidth]{gfx/ConstClust/AnalisisImagenes/Aibo3}} \quad
	\caption[Clustering method used in Aibo robot navigation system.]{Clustering method used in Aibo robot navigation system \cite{davidson2007survey,davidson2005clustering}.}\label{fig:AiboRobotClustSys}
\end{figure}

@@

Video databases are one example where constraints can be generated directly from the data domain, especially if space-time data information is available \cite{yan2006discriminative}. In time sequenced data it is possible to set \acs{ML} constraints between groups of pixels of frames close in time. This is especially useful when the task is to perform object recognition based on clustering and segmentation. It is also possible to add \acs{CL} constraints to image segment in the same snapshot, as there is a low probability that they are associated with the same object after segmentation is done. In fact, in the video analysis domain problems there are a variety of constraint extraction methods \cite{yan2006discriminative}, Figure \ref{fig:ConstExtractionVideo} shows some examples.

\begin{figure}[bth]
	\myfloatalign
	\subfloat[]
	{\includegraphics[width=.4\linewidth]{gfx/ConstClust/Videos/VideoA}
	\label{fig:ConstExtractionVideoA}}
	\quad
	\subfloat[]
	{\includegraphics[width=.225\linewidth]{gfx/ConstClust/Videos/VideoB}
	\label{fig:ConstExtractionVideoB}} \quad
	\subfloat[]
	{\includegraphics[width=.4\linewidth]{gfx/ConstClust/Videos/VideoC}
	\label{fig:ConstExtractionVideoC}}
	\quad
	\subfloat[]
	{\includegraphics[width=.4\linewidth]{gfx/ConstClust/Videos/VideoD}
	\label{fig:ConstExtractionVideoD}}
	\caption[Constraints extraction in video data.]{Constraints extraction in video data \cite{yan2006discriminative,davidson2007survey}.}\label{fig:ConstExtractionVideo}
\end{figure}

In Figure \ref{fig:ConstExtractionVideo}, image \ref{fig:ConstExtractionVideoA} corresponds to constraints extracted from tracking a person over a period of time, \ref{fig:ConstExtractionVideoB} corresponds to spatial constraints associating two objects located in the same frame, image \ref{fig:ConstExtractionVideoC} corresponds to restrictions obtained through facial recognition and \ref{fig:ConstExtractionVideoD} to those provided by the user.

@@

In gene clustering based on micro-array data, genes are represented by their expression profile in different experiments and grouped using different constrained clustering methods. Figure \ref{fig:GeneticApp} shows an example: these are \acs{ML} constraints set between genes based on co-occurrence data stored in the protein interaction database, which contains information about which genes (and their associated proteins) are present in the same cell processes \cite{xenarios2001dip}. This information can be used to improve results provided by clustering methods. \cite{segal2003discovering}.

\begin{figure}[!h]
	\centering
	\includegraphics[scale=0.3]{gfx/ConstClust/Genetica/Genes} 
	\caption[Gene clustering based on micro-array data.]{Gene clustering based on micro-array data \cite{davidson2007survey}.}\label{fig:GeneticApp}
\end{figure}

@@

In content classification tasks, the goal is to automatically split large amounts of documents into groups or clusters. In this case it is possible to extract constraints from multiple auxiliary resources. For example, if two documents are in the same directory, a \acs{ML} constraints could be set between them. This way it is possible to adjust the resulting clustering to suit a particular criterion, such as creating a hierarchy of documents similar to the way they are organized in the input directories structure.

@@

Constrained clustering is very useful in web search data processing. Here, the aim is to automatically group the results of an ambiguous query in the search engine into clusters of \acs{URL}s, which refer to the concept introduced as a query in different contexts. In this area it is possible to extract the constraints from searches previously performed by users, so that a \acs{ML} constraints is set between \acs{URL}s visited in the same user session. Clustering using these constraints can help bias the result of searches towards user's preferences.

\subsection{Audio Data}

Constrained clustering is very useful in the processing of search data on web pages. Here, the aim is to automatically group the results of an ambiguous query in the search engine into clusters of \acs{URL}s, which refer to the concept introduced as a query in different contexts. In this area it is possible to extract constraints from searches previously performed by users, so that a \acs{ML} constraint is set between visited \acs{URL}s in the same user session. Clustering using these constraints can help biasing search results towards user preferences.

\subsection{GPS Data} \label{sec:GPSApp}

@@

\begin{figure}[!h]
	\centering
	\includegraphics[scale=0.3]{gfx/ConstClust/GPS/Coches} 
	\caption[GPS data use.]{\acs{GPS} data use. \cite{davidson2007survey} \cite{wagstaff2001constrained}}\label{fig:GPSClustData}
\end{figure}

In this domain, real clusters have an elongated shape on the horizontal axis and are aligned perpendicularly to the motion direction. To get the resulting clusters to have this shape, we can make use of the constraints. We will set a \acs{CL} constraint between those instances more than four meters apart perpendicularly to the motion direction (since the lanes have a maximum width of four meters), and \acs{ML} constraints between those instances that present continuity in the motion direction axis, since it is likely that the vehicles they represent are in the same lane. This clustering model has proven to be very useful in real-time navigation \cite{wagstaff2001constrained}, allowing the user to be notified when to change lanes, or when not to leave them.

\begin{figure}[!h]
	\centering
	\includegraphics[scale=0.32]{gfx/ConstClust/GPS/Instancias} 
	\caption[Clustering of GPS data with no constraints.]{Clustering of \acs{GPS} data with no constraints \cite{davidson2007survey,wagstaff2001constrained}.}\label{fig:figure15}
\end{figure}

@@

The already mentioned constrained clustering applications are the most studied in the literature, although the reader may have the impression that constrained clustering is no longer used in real-world problems. To prevent that we provide a list of some more recent applications of constrained clustering to modern problems: advanced robotics navigation systems \cite{semnani2016constrained}, applied marketing \cite{seret2014new}, obstructive sleep apnea analysis \cite{mai2018evolutionary}, handwritten digits classification \cite{li2015scalable}, Internet traffic classification \cite{wang2014internet}, electoral district designing, \cite{brieden2017constrained}, semantically similar relations identification \cite{wang2015constrained}, cooperative team building \cite{yang2014team}, video face clustering \cite{zhou2014video}, image clustering \cite{habashi2017semi} and storage location assignment in warehouses \cite{yang2016constrained} among others.

@@

Given $X = \{x_1 \cdots x_n\}$, a large set of unlabeled instances, and $L = \{(x_{n+1}, l_{n+1}) \cdots (x_{n+|L|}, l_{n+|L|})\}$, a small set of labeled instances, it is common to choose two elements from $L$ (with replacement) and set a \acs{ML} constraint between them if they belong to the same class, or a \acs{CL} constraints otherwise. An appropriate method to evaluate the results provided by a clustering method is to measure the level of accuracy it achieves when predicting dataset $X$ labels. This normally requires specifying the number of desired clusters equal to the number of known classes in $X$ ($k = k^*$). Methods such as the RandIndex \cite{rand1971objective} or its variants, which will be discussed later, are used to measure the degree of agreement between two given partitions (clusterings).

In \cite{wagstaff2000clustering}, where constraints are generated as described above, it is shown that accuracy is increased up to a 20\% over classic methods when averaging results over different constraint sets.

@@

	\textbf{Using constraints increases average accuracy.}
	The performance of a method in predicting labels increases when averaged using numerous different constraint sets \cite{davidson2007survey}.
	\label{ob:AccuracyIncrease}

@@

This rule, however, is not true in all cases, as in datasets such as \textit{Tic-Tac-Toe Endgame}, no increase in predictions is achieved regardless of the number of constraints used. A possible explanation for these exceptions is based on the fact that setting $k = k^*$ may not be appropriate in these cases.

@@

Although, as we have seen, incorporating constraints to clustering methods brings benefits in some applications, there are two main drawbacks that are discussed below, as well as possible solutions to them.

@@

Introducing constraints in clustering changes the problem it solves, which becomes: \textit{Find the best partition that satisfies all constraints}. This way, if constraints are not properly specified or if the extraction methods are inadequate, we can find that the constraints contradict each other, which means that there is no partition that satisfies them all. For instance, there is no partition that satisfies the constraints $C_=(x_i,x_j)$ and $C_{\neq}(x_i,x_j)$, regardless of the value of $k$. The same happens for $k = 2$, and the constraints $C_{\neq}(x_i, x_j)$, $C_{\neq}(x_j, x_q)$ and $C_{\neq}(x_i, x_q)$. Definition \ref{def:FeasibilityProblem} formalizes the feasibility problem for non-hierarchical instance-level constrained clustering:

@@

	\textbf{Feasibility problem for non-hierarchical instance-level constrained clustering:} Given a dataset $X$, a constraint set $CS$, and the bounds on the number of clusters $k_l \leq k \leq k_u$, does there exist a partition $C$ of $X$ with $k$ clusters such that all constraints in $CS$ are satisfied? \cite{davidson2005clustering,davidson2007survey}

@@

	\textbf{Constrained clustering with \acs{CL} is analogous to the graph coloring problem} \cite{davidson2007survey}.
